---
title: South Pacific invasive alien species gap analysis in DNA barcode reference libraries
date: "`r Sys.Date()`"
author: Pieter Provoost
output: rmdformats::robobook
editor_options: 
  chunk_output_type: console
---

## Introduction

The goal of this analysis is to find out if there are any gaps in sequence reference databases for species that are of interest to the PacMAN project.

This notebook makes use of the [caspr](https://github.com/iobis/caspr) package, which is a set of R functions created to support data analysis within the PacMAN project.

## Region of interest

```{r message=FALSE, warning=FALSE}
library(caspr)
library(sf)
library(dplyr)

library(leaflet)

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = st_as_sfc(wkt_south_pacific, crs = 4326), color = "#ff9900")
```

## Creating the species checklist

Let's first get the expert priority species lists, which are [managed by the PacMAN team on GitHub](https://github.com/iobis/pacman/tree/main/SpeciesLists), as well as a checklist for the South Pacific based on the OBIS database and the World Register of Introduced Marine Species (WRiMS). The checklists are merged to obtain a single checklist.

```{r message=FALSE}
cl_expert <- expert_checklists(enhance = TRUE) %>%
  select(taxonID, scientificName, taxonRank, phylum, references, remarks)
cl_obis <- obis_checklist()

cl <- cl_obis %>%
  full_join(cl_expert) %>%
  filter(taxonRank == "Species")
```

This gives us a checklist with `r nrow(cl)` species from `r length(unique(cl$phylum))` phyla. Of these species, `r nrow(cl_expert %>% filter(taxonRank == "Species"))` are on the expert priority species lists.

```{r warning=FALSE, message=FALSE}
library(ggplot2)

phylum_stats <- cl %>%
  group_by(phylum) %>%
  summarize(species = n()) %>%
  arrange(desc(species))

ggplot() +
  geom_histogram(data = phylum_stats, aes(x = reorder(phylum, -species), y = species), stat = "identity") +
  xlab("phylum") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Checking BOLD for sequences

Of the `r nrow(cl)` species in our checklist, `r sum(!is.na(cl$bold_id))` have a BOLD identifier in the WoRMS database, but in this first implementation we are querying BOLD by taxon name and not by identifier.

Let's get the number of sequences in BOLD by species and marker:

```{r}
library(bold)
library(purrr)
library(progress)

if (file.exists("df_bold.rds")) {
  df_bold <- readRDS("df_bold.rds")
} else {
  bold_list <- sapply(cl$species, function(x) NULL)
  pb <- progress_estimated(nrow(cl))
  for (i in 1:length(cl$species)) {
    if (is.null(bold_list[[i]])) {
      taxon <- cl$species[i]
      bold_list[[i]] <- caspr::bold_statistics(taxon)  
    }
    pb$tick()$print()
  }
  df_bold <- bind_rows(bold_list, .id = "id")
  saveRDS(df_bold, "df_bold.rds")
}
```

This gives us a table with `r length(unique(df_bold$markercode))` distinct marker codes and `r length(unique(df_bold$id))` species.

```{r warning=FALSE, message=FALSE}
marker_stats <- df_bold %>%
  group_by(markercode) %>%
  summarize(species = n(), sequences = sum(sequences))

ggplot() +
  geom_histogram(data = marker_stats, aes(x = reorder(markercode, -species), y = species), stat = "identity") +
  xlab("marker") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

Let's take a look at the completeness by phylum. For this analysis I'm verifying that for each taxon at least one of the following markers is present: COI, 18S, rbcL, matK ^[Duarte S, Vieira PE, Costa FO (2020) Assessment of species gaps in DNA barcode libraries of non-indigenous species (NIS) occurring in European coastal regions. Metabarcoding and Metagenomics 4: e55162. https://doi.org/10.3897/mbmg.4.55162].

```{r warning=FALSE, message=FALSE}
selected_markers <- c("COI-5P", "18S", "rbcL", "rbcLa", "matK")

df_bold <- cl %>%
  left_join(df_bold, by = c("scientificName" = "id")) %>%
  mutate(has_barcode = markercode %in% selected_markers)

species_barcode_stats <- df_bold %>%
  group_by(phylum, species, records) %>%
  summarize(has_barcode = as.logical(max(has_barcode)))

phylum_barcode_stats <- species_barcode_stats %>%
  group_by(phylum, has_barcode) %>%
  summarize(species = n()) %>%
  arrange(desc(has_barcode))

ggplot() +
  geom_histogram(data = phylum_barcode_stats, aes(x = phylum, y = species, fill = has_barcode), stat = "identity") +
  scale_fill_grey(start = 0.7, end = 0.3) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

Let's check which species without barcodes have the most occurrences in OBIS. This table gives the five most abundant species for each phylum:

```{r message=FALSE, warning=FALSE}
library(knitr)

missing_barcodes <- species_barcode_stats %>%
  filter(has_barcode == FALSE)

missing_barcodes %>%
  arrange(phylum, desc(records)) %>%
  group_by(phylum) %>%
  slice_head(n = 5) %>%
  kable()
```

## Checking NCBI for sequences

To do.

## Issues to be resolved

- limit checklist to species introduced in the region
- query BOLD by taxon ID if available
- NCBI
- create priority list for barcoding specimens
